Internal models of type theory

### 2013-12-09 article-models-intro.tex
working on the introduction

\section{Introduction}
This project has its origins in the beginning of 2013, when I proved a version
of the descent theorem for homotopy colimits in type theory while I was working
with Bas Spitters to develop notions from higher category theory in the
univalent foundations. To arrive at a
notion of diagram general enough to capture the higher inductive types described
in chapter 6 of \cite{TheBook} excluding the truncations we needed type
theoretical graphs. The graphs form a model of type theory and indeed we needed
several of the basic type theoretical operations to give an efficient approach
to the descent theorem. However it was not an issue to describe the graph model
and the sense in which it was a model, the presentation of it was unsystematic
and in that sense unsatisfactory too.

From the experience of working with the graph model it became appearent that the
answer we had of the question of what it models, i.e.~Martin L\"of type theory
with the univalence axiom, could have been different if we were to describe a
type theory based on the mechanism on which the model was built. The particular
featueres of the graph model were that there was a type of graphs resembling a category
of contexts, for each context a notion of what it means to be a family of graphs
and a term thereof; there are the operations of context extension, weakening and
substitution. In other words, the graph model is an instance of type theretical
categories with families, an invention of Peter Dybjer.

As we know however, the type theoretical description of a category as given in
chapter 11 of \cite{TheBook} is restricted to categories which have a $1$-type
as the type of objects and for every two objects a $0$-type of morphisms.
Although categories with families are like that too, these are not very natural
restrictions when one is concerned about internal models of type theory. It
seems we're in the need of a higher categorical formulation of what an internal
model, or a category with family, is. But type theory has the tendency to take
care of it's own higher categorical structure and one could hope that type
theory is an elementary theory of the category with families of category with
families.

>>>>

\section{Introduction}
The project I propose here has its origins in the beginning of 2013, when I proved a version
of the descent theorem for homotopy colimits in type theory while I was working
with Bas Spitters to develop notions from higher category theory in the
univalent foundations. To arrive at a
notion of diagram general enough to capture the higher inductive types described
in chapter 6 of \cite{TheBook} excluding the truncations we needed type
theoretical graphs. The graphs form a model of type theory and indeed we needed
several of the basic type theoretical operations to give an efficient approach
to the descent theorem. Although it was not an issue to describe the graph model
and the sense in which it models the type constructors, 
not all of type theory is interpreted very well. To start with, context
extension isn't strictly associative if dependent pair types are used for
its interpretation. Also, if $A$ and $B$ are families of graphs
over a graph $\Gamma$, then $B$ isn't also a family of graphs over the extended
graph $\ctxext{\Gamma}{A}$ (interfering with a good interpretation of weakening).
There are no issues with the interpretation of substitution though. Actually
it is not so straightforward to describe what it models!

The graph model
shows three operations: extension, weakening and substitution. All of them
can be defined such that they act not only on contexts, but also on families
and terms. Moreover, they are all compatible with each other. In fact, along
with asserting the existence of identity morphisms (which is done in type
theory), it seems
that this structure is enough to establish some of the structure of a category.
We might even go so far as to boldly assert that
\begin{quote}
\emph{Category theory is dependent type theory without type constructors.}
\end{quote}
It is a part of this proposal to test this hypothesis, with the note that by
`category theory' we mean a theory of weak higher categories which does not
have any of the truncatedness restrictions -- in particular not those appearing
in the theory of AKS-categories as
described in \cite{TheBook} in chapter 11. We note that even Dybjers theory
of internal categories with families has such restrictions: he relies on setoids.
It needs not much arguing that these are very unnatural from a univalent point
of view.

### 2013-12-09 article-models-tt.tex
working on the introduction

\subsection{The identity functions}
Without a rule explicitly asserting the existence of an identity morphism we don't
get one, hence we do that here. The identity morphism is a term which introduced
in ordinary type theory via the variable rule. The variable rule is a bit more
general: it asserts that
\begin{equation*}
\jterm{\Gamma,\,x_1:A_1,\ldots,\,x_n:A_n}{A_i}{x_i}
\end{equation*}
for every $1\leq i\leq n$. Thus, it establishes the projections. In our setting,
we get the projections from the identity morphisms together with weakening. We
already have weakening, so here it suffices to introduce the identity morphisms.
They are determined by their behavior with respect to substitution combined with
weakening.

@@

\subsection{Stability under extension}
We need to come up with a suitable notion of stability under extension, which
is to be the type theoretical analogue of category theoretical stability under
slicing. When we have done this, we have to show that the constructions we have
so far are indeed stable under extension.

We can make weakening stable under extension by the following
process. Suppose we have a context $\Gamma$. For a context $\Delta$ we add the
judgment
\begin{equation*}
\vdash \Delta~\tfwk(\Gamma)-admissible
\end{equation*}
for the assertion that $\Delta$ is allowed as a subject for the action of
$\tfwk(\Gamma)$. If $\Delta$ is a $\tfwk(\Gamma)$ admissible context and

>>>>

\subsection{Terms in the basic type theory}

\subsubsection{Identity functions}
Without a rule explicitly asserting the existence of an identity morphism we don't
get one, hence we do that here. The identity morphism is a term which introduced
in ordinary type theory via the variable rule. The variable rule is a bit more
general: it asserts that
\begin{equation*}
\jterm{\Gamma,\,x_1:A_1,\ldots,\,x_n:A_n}{A_i}{x_i}
\end{equation*}
for every $1\leq i\leq n$. Thus, it establishes the projections. In our setting,
we get the projections from the identity morphisms together with weakening. We
already have weakening, so here it suffices to introduce the identity morphisms.
They are determined by their behavior with respect to substitution combined with
weakening.

@@

The identity functions in combination with weakening give rise to the projections.
We define

\begin{equation*}
\proj A \defeq \ctxwk{A}{\idfunc[\Gamma]}:\ctxwk{A}{{\Gamma}{\Gamma}}.
\end{equation*}

\subsubsection{Inclusions}
In ordinary type theory, there is a rule
\begin{equation*}
\inference{\jterm{\Gamma}{A}{x}\qquad\jtype{\Gamma,\,a:A}{P(a)}}
  {\Gamma,\,u:P(x/a)\vdash x:A, u:P(x/a)}
\end{equation*}
This rule gives the inclusion of the fiber $\subst{x}{P}$ into $\ctxext{A}{P}$.

Thus, we require
\begin{equation}
\inference{\jterm{\Gamma}{A}{x}\qquad\jtype{\ctxext{\Gamma}{A}}{P}}
  {\jterm{\ctxext{\Gamma}{\subst{x}{P}}}{\ctxwk{\subst{x}{P}}{\ctxext{A}{P}}}{\typefont{in}^P_x}}
\end{equation}

The terms $\typefont{in}^P_x$ have the following behavior:

\begin{align}
& \inference
    {\jtype{\ctxext{{\Gamma}{A}}{P}}{Q}}
    {\jtypeeq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{\typefont{in}^P_x}{\ctxwk{\subst{x}{P}}{Q}}}{\subst{x}{Q}}}\\
& \inference
    {\jterm{\ctxext{{\Gamma}{A}}{P}}{Q}{g}}
    {\jtermeq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{\typefont{in}^P_x}{\ctxwk{\subst{x}{P}}{g}}}{\subst{x}{g}}}
\end{align}
